DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-12-12 2015CompleteDerivation/completeDerivation.tex
Incorporate discussion with Dave and Lydia on state counting

We'll see that very similar relationships are what define Hamiltonian and Lagrangian mechanics. But recovering them in the continuous case is trickier, and in fact will be part of the challenge in the following sections. Consider the map $q'=aq$ with $0<a<1$. At first glance, it's a bijective continuous map so we may think it can represent a deterministic and reversible evolution. Yet, $\Delta q \equiv [-b, b] \supset \Delta q'$: a set of states is mapped to a proper subset (i.e. to fewer states) which does not make sense for a reversible process. In the limit where we apply the map an infinite amount of times, any value of $q$ will be brought infinitely close to $0$, which also does not sound like a reversible process. The issue is that while deterministic evolution will map points to points, not all point to point maps can be considered deterministic and reversible. Discrete sets already come with a way to measure and compare the number of elements which is preserved by any bijective map. With continuous sets, we need to do extra work to properly define a measure (or metric) that allows one to count states and possibilities.

@@

As we can see, the framework we have is ill suited to handle these cases. Therefore we cannot simply stick time in the Hamiltonian and expect everything to work out. But we can't ignore the problem either as the three cases outlined are pretty common situations to study. So we need to go back to our definitions and amend them properly.

>>>>

We'll see that very similar relationships are what define Hamiltonian and Lagrangian mechanics. But recovering them in the continuous case is not as straight forward. Consider the map $q'=aq$ with $0<a<1$. At first glance, it's a bijective continuous map so we may think it can represent a deterministic and reversible evolution. Yet, $\Delta q \equiv [-b, b] \supset \Delta q'$: a set of states is mapped to a proper subset (i.e. to fewer states) which does not make sense for a reversible process. In the limit where we apply the map an infinite amount of times, any value of $q$ will be brought infinitely close to $0$, which also does not sound like a reversible process.

The issue is that while deterministic evolution will map states to states, not all bijective state to state maps can be considered deterministic and reversible. This is a crucial point where this work is different from more usual treatments of dynamical systems. If one focuses only on the states themselves and counts them, it reaches the conclusion that a range of $1$m contains the same number of initial conditions of $1$Km: they have the same infinite number of elements. This conclusion has a number of problems: it does not match our physical intuition; it's not consistent with all types of topology; it does not explain why the laws of evolution are very often, not only continuous, but differentiable. In short: a bijection does not capture the entirety of the physical description.

If we work with a discrete topology then, by all means, the way to measure the size of a set is to count the elements in a set. But if we use the standard topology on $\mathbbm{R}$, isolated points have measure zero: what defines the number of initial conditions is the range of the state variables. As we'll see, the geometrical structures in Hamiltonian/Lagrangian classical/quantum mechanics are there to allow us to properly count states and, more specifically, possibilities along independent state variables. If we think that counting states is equivalent to counting elements in the set, we miss the whole point of that geometrical structure.

What makes the matter even more confusing is that the single physical idea of counting states and possibilities along independent state variable does not map to a single mathematical concept. Sometimes we need a measure, sometimes we need a symplectic form and sometimes we need a metric tensor. It will depend, case by case, on the state space of the system under study. But, however defined mathematically, deterministic and reversible motion must preserve that count. This is the crucial insight that gives physical meaning to those structures, that justifies the use of differential equations for state evolution, that tells us why symplectic and Riemannian geometry are so important in physics. Properly characterizing this count will be one of the key goals of the following sections.

@@

As we can see, the framework we have is ill suited to handle these cases. Therefore we cannot simply stick time in the Hamiltonian and expect everything to work out.\footnote{Looking ahead, another way to frame the problem is that Galilean transformations do not preserve phase space volume. Lorentz transformations, on the other hand, do and it is this requirements plus the kinematic assumption that will recover them later.} But we can't ignore the problem either as the three cases outlined are pretty common situations to study. So we need to go back to our definitions and amend them properly.
